Lattice Boltzmann study of mass transfer for complicated geometries

### 2012-01-21 Text/introduction.tex
Added some average calculations.

\newcommand{\beq}{\begin{equation}}
\newcommand{\feq}{\end{equation}}
\newcommand{\beqstar}{\begin{equation*}}
\newcommand{\feqstar}{\end{equation*}}                                     
\newcommand{\beqal}{\begin{equation}\begin{aligned}}
\newcommand{\feqal}{\end{aligned}\end{equation}}
\newcommand{\beqalstar}{\begin{equation*}\begin{aligned}}
\newcommand{\feqalstar}{\end{aligned}\end{equation*}}
\newcommand{\vol}{k_L\,a}

@@

\begin{document}
\maketitle
\section{Mass transfer definitions}
By the definition the mass transfer coefficient is the following:
\beq
k_L=\frac{\dot{m}}{P \Delta C},
\feq
where $\dot{m}$ is the mass rate $\Bigl[\frac{kg}{s}\Bigr]$, $P$ is the area of the surface
$\Bigl[m^2\Bigr]$, $\Delta C$ is the concentration difference $\Bigl[\frac{kg}{m^3}\Bigr]$.
Therefore, $k_L$ has a dimension of velocity $\Bigl[\frac{m}{s}\Bigr]$.

There are different methods to estimate the mass transfer coefficient $k_L$. We first examine the
theoretical definitions of the mass transfer in case of the point sources.
\subsection{Point mass sources}
In what follows we will present three approaches to calculate the mass points mass transfer
coefficients:
\begin{enumerate}
\item
Let us look at the infinitisemal small domain of the volume $A \Delta x$ in a steady-state, not
moving and with the mass source. Then the concentration difference can be found as $\Delta C = C^* -
C(t)$, where $C^*$ is the imposed concentration on the infinitesimal concentration source, $C(t)$ is
the time depending average concentration. Therefore one can write the time dependent ODE for the
average concentration in the domain:
\beq
\dot{m}= A \Delta x \frac{\mathrm{d}C}{\mathrm{d} t} = k_L P (C^{*}-C(t)), 
\feq
with the initial condition $C(0)=0$
The solution in time becomes as follows:
\beq
C(t)= C^{*}(1-\exp(-k_L a t )), 
\feq
where $k_L a$ is the volumetric mass transfer coefficient.

\item
The previous consideration is able to predict mass transfer happening in moving with the velocity
$U$ liquid, see Fig. \ref{fig:moving:frame}.
\begin{figure}[htb!]
\includegraphics[width=\textwidth]{Figures/mass_transfer.eps}
\caption{The mass transfer in the moving liquid. \label{fig:moving:frame}}
\end{figure}

@@

\subsection{Numerical calculation}
In the numerical simulation it is not possible to simulate really long domains with the infinite
number of the bubbles. As we saw before there are two approaches towards it - either to simulate
the bubble train and then to measure concentration along the pipe, Eq.
\ref{main:mass:transfer:expression}, or to transfer to the reference frame moving with bulk
velocity $U$ and conduct the same measurements through Eq. \ref{main:mass:transfer:expression}.
However, both methods do require a tracking of moving bubbles which is complicated from the
numerical point of view.

@@

>>>>

\newcommand{\beq}{\begin{equation}}
\newcommand{\feq}{\end{equation}}
\newcommand{\beqstar}{\begin{equation*}}
\newcommand{\feqstar}{\end{equation*}}                                     
\newcommand{\beqal}{\begin{equation}\begin{aligned}}
\newcommand{\feqal}{\end{aligned}\end{equation}}
\newcommand{\beqalstar}{\begin{equation*}\begin{aligned}}
\newcommand{\feqalstar}{\end{aligned}\end{equation*}}
\newcommand{\vol}{k_L\,a}
\newcommand{\lbubble}{L_{\mathrm{bubble}}}
\newcommand{\lunit}{L_{\mathrm{unit}}}
\newcommand{\lslug}{L_{\mathrm{slug}}}
\newcommand{\ububble}{U_{\mathrm{bubble}}}
\newcommand{\uliq}{U_{\mathrm{liq}}}
\newcommand{\ugas}{U_{\mathrm{gas}}}
\newcommand{\cbubble}{C_{\mathrm{bubble}}}

@@

\begin{document}
\maketitle
\begin{abstract}
This work gives the thorough procedure of how to determine the volumetric mass transfer coefficient
in the context of lattice Boltzmann simulations. The novelty of this work lies in the establishing
a thorough procedure of how to measure the volumetric mass transfer coefficient in case of mass
sources to be comparable of the representative unit cell, which accounts for the inhomogenous
situation. By examining $5$ representative cases taken from microfluidics with the capillary number
from $Ca=0.1$ to $Ca=1$ the following relationship was established {\color{red} Give main findings
here}. The work is of use to people performing mass transfer numerical simulations.
\end{abstract}

\section{Introduction}
The volumetric mass transfer coefficient is a very important characteristic for microchannels 
design \cite{kreutzer-overview}. It allows to
properly manufacture a microchannel with necessary properties to ensure that chemical
reactions are performed in the best manner.

The volumetric mass transfer coefficient correlations are represented as a function of the diffusion
coefficient, slug and bubble lengths, and bubble velocity. For example, \citet{yue-mass} established
experimental correlation for periodic bubble train as: 
\begin{equation}
\vol =\frac{2}{d_h} \Bigl(\frac{D
\ububble}{\lbubble+\lslug}\Bigr)^{0.5}
\Bigl(\frac{\lbubble}{\lbubble+\lslug}\Bigr)^{0.3},
\end{equation}
where $\vol$ is the volumetric mass transfer coefficient, $d_h$ is the hydrodynamic capillary
diameter, $\lbubble$ is the bubble length, $\lslug$ is the slug distance (between bubbles),
$\ububble$ is the bubble velocity. This correlation is within $10\%$ accuracy for experimental
results for the following range of
parameters: bubble velocity $0.4\,\mathrm{m/s}<\ububble<2\,\mathrm{m/s}$, bubble length
$1.4<\frac{\lbubble}{d_h}<6.3$, slug length $1<\frac{\lslug}{d_h}<3.2$,
hydrodynamic diameter $d_h = 400 \mathrm{\mu m}$.

Other correlations can be found in works of \citet{kreutzer-overview}
and \citet{bercic-mass}. Among a variety of experimental works extensively covered
\cite{yue-mass}, the numerical validations to the best author knowledge are sparse. One of the
reasons is that the experimental correlations are established for continous system, i.e. one does
not distinguish separate bubbles, but looks to the system as a whole measuring mass concentration
in different places. There are some works as how to avoid this approach and perform numerical
simulations for one unit cell. For example, \citet{vanbaten-circular} perform numerical simulations
for one unit cell using periodic boundary conditions and measure the volumetric mass transfer
coefficient as:
\begin{equation}
\label{main:simulation:equation}
k_L a=\frac{\text{Flux}}{\cbubble-C_{outlet}} \frac{\text{bubble surface area}}{\text{unit cell
volume}},
\end{equation}
where $C_{outlet}=\int{C U_{outlet}\mathrm{d}A}/\int{U_{outlet}\mathrm{d}A}$. 
The concentration flux is calculated as the difference between overall
average concentration taken in the whole domain ($C_{overall}=\int_{V} C \mathrm{d}V /V$)
at time
$t_1$ and time $t_2$ divided on the time difference $t_2-t_1$. The agreement between numerical
simulations and correlations of \citet{bercic-mass} was good. However, the numerical model only
accounts for circular capillaries and short contact times (the mass t), which indicate that the
bubble film is not saturated with the mass concentration. As well, authors took the approximation of
bubble shapes for small capillary numbers $Ca<0.01$ and then derived the hydrodynamics fields by
imposing a free-slip condition on the bubble surface.

This work aims at the establishing numerical procedures as to how properly obtain the volumetric
mass transfer coefficient in the interesting range of capillary numbers, i.e. $0.1\leq Ca\leq1.0$.
The comparison of volumetric mass transfer coefficient correlations with experimental correlations
and establishing the Sherwood number as a function of other non-dimensional parameters is the goal
of the future work.

We chose this type of the problem due to different reasons:
\begin{description}
\item[I] The gas holdup, i.e. the ratio of volume occupied by bubble to unit cell volume, is
significant. Therefore, one cannot assume that the mass source is a point source or one can have
homogenous picture. Therefore, to establish the volumetric mass transfer coefficient one needs to
mimic the continuos picture. One of the goals of this work is to establish certain criteria as how
to simulate it numerically.
\item[II] For small capillary numbers as $Ca<0.7$ \cite{giavedoni-numerical} there is a vortex in
the reference frame moving with
the bubble, which  influences and improves the corresponding mass transfer.  For larger capillary
numbers there is no vortex and stremalines go around the bubble. Thus, the problem is interesting
because of really different hydrodynamic patterns depending on capillary number. Note that the flow
patterns strongly depend on the number of parameters as the slug length, the thickness of the
thickness and the existence of the vortex in front of the bubble. There are some available
correlations in literature \cite{bercic-mass,kreutzer-overview} and one of the particular goal of
this work is to study mass transfer dependence on the number of parameters. 
\end{description}

The approach we want to use is the lattice Boltzmann method, a relatively new CFD
competitor emerged during last $20$ years \cite{frisch,mcnamara,HJ,HSB}. During years the method
was applied to simulate not only hydrodynamic problems \cite{yu}, but as well multiphase flows
\cite{Shan-chen:extended,swift,gunstensen}, heat transfer
\cite{yuan-thermal,zhang-thermal}, ferrofluids \cite{dellar-ferro,kuzmin-aniso}.

The mass transfer problems in the lattice Boltzmann framework were mainly addressed in series of
works of Dr. Ginzburg and co-authors
\cite{ginzburg-main,ginzburg-boundary-conditions,ginzburg-saturated-flow}. However, all these works
are of general nature to simulate the advection-diffusion equation via the lattice Boltzmann
framework. The manuscripts do not establish of how to obtain the volumetric mass transfer
coefficients. The work of \citet{inamuro-scalar-boundary} simulates heat and mass transfers in the
porous media. The work of \citet{jos-mass} simulates lateral mixing in cross-channel flow. While
two last works are focused at the mass transfer problems, both problems are of homogenuous nature
and do not guide as to how to simulate the non-homogenous problems.

The overall goal of this paper is to address the questions arised and to establish the numerical
procedure suitable to measure the volumetric mass transfer coefficient. We will utilize the results
of numerical simulations of flow
obtained
before \cite{kuzmin-binary2d}.

The paper is organized as follows. The mass transfer coefficients and volumetric mass transfer
coefficients are covered. The limiting cases from the continuos representation and their
application to the present simulation are given. The mass transfer simulations benchmarks are
presented. Finally, numerical simulations of hypotheses presented in this work and comparison
between them are presented.

\section{Mass transfer definitions}
By the definition the mass transfer coefficient from the surface with the imposed constant
concentration $\cbubble$ is the following:
\beq
k_L=\frac{\dot{m}}{P \Delta C},
\feq
where $\dot{m}$ is the mass rate $\Bigl[\frac{kg}{s}\Bigr]$, $P$ is the area of the surface
$\Bigl[m^2\Bigr]$, $\Delta C$ is the concentration difference between the surface and the medium
$\Bigl[\frac{kg}{m^3}\Bigr]$. Therefore, $k_L$ has a dimension of velocity
$\Bigl[\frac{m}{s}\Bigr]$. Usually, the medium concentration is taken at the infinite distance
from the bubble. However, in the case bubble, the medium concentration can be the average
concentration in the domain, the flux averaged concentration at the outlet. Thus, one needs to
establish a thorough definition of the volumetric mass transfer coefficient in the case of complex
geometries and hydrodynamic patterns.

There are different methods to estimate the mass transfer coefficient $k_L$. We first examine the
theoretical definitions of the mass transfer in case of the point sources.
\subsection{Point mass sources}
In what follows we will present three approaches to calculate point mass transfer
coefficients:
\begin{enumerate}
\item
Let us look at the infinitisemal small domain of the volume $A \Delta x$ in a steady-state, not
moving and with the mass source. Then the concentration difference can be found as $\Delta C = C^* -
C(t)$, where $C^*$ is the imposed point source concentration, $C(t)$ is
the time depending average concentration. Therefore one can write the time dependent ODE for the
average concentration in the domain:
\beq
\dot{m}= A \Delta x \frac{\mathrm{d}C}{\mathrm{d} t} = k_L P (C^{*}-C(t)), 
\feq
with the initial condition $C(0)=0$
The solution in time becomes as follows:
\beq
C(t)= C^{*}(1-\exp(-\vol t )), 
\feq
where $\vol$ is the volumetric mass transfer coefficient defined as:
\beq
\vol=k_L \frac{P}{V},
\feq
where $P$ is the source surface, $V$ is the unit cell volume.
\item
The previous consideration is able to predict mass transfer happening in moving with the velocity
$U$ liquid, see Fig. \ref{fig:moving:frame}.
\begin{figure}[htb!]
\includegraphics[width=\textwidth]{Figures/mass_transfer.eps}
\caption{The mass transfer in the moving liquid. \label{fig:moving:frame}}
\end{figure}

@@

\subsection{Numerical calculation}

The numerical benchmark is presented in Fig. \ref{fig:benchmark}. In
the same figure, one can see the proposed benchmark for the tracer. The
physical
picture for the proposed numerical benchmark is  related to classical experiments conducted by
\citet{bercic-mass}. In their work authors studied gas-liquid mass transfer coefficient by
injecting methane as the gas bubble and measuring the dissolved concentration of methane in water
along the channel. This situation implies the periodic boundary conditions and steady-state motion
of bubbles.

The numerical  procedure of studying mass transfer is presented in the following steps:
\begin{description}
 \item[Flow field] The hydrodynamics for the bubble motion in the periodic domain is obtained
according to the work \cite{kuzmin-binary2d}. The grid resolution is taken to insure grid
independency of results.  
 \item[Bubble reference frame] Once the hydrodynamics is resolved, the mass transfer simulations
are conducted in the moving
reference frame with the bubble. Then the bubble stands still and the flow is coming around the
bubble. We impose a steady concentration of solutant on the surface of the bubble.
\item[Mass transfer] One needs not only perform a Galilean transformation to the reference
frame moving with the bubble but also to
transfer the bubble to the inlet condition. This can insure that the studied phenomenon is a liquid
slug. 
\end{description}

In the numerical simulation it is not possible to simulate really long domains with the infinite
number of the bubbles. As we saw before there are two approaches towards it - either to simulate
the bubble train and then to measure concentration along the pipe, Eq.
\ref{main:mass:transfer:expression}, or to transfer to the reference frame moving with bulk
velocity $U$ and conduct the same measurements through Eq. \ref{main:mass:transfer:expression}.
However, both methods do require a tracking of moving bubbles which is complicated from the
numerical point of view.

@@

\section{Performing average approach}
One can calculate the average domain concentration over the time. According to the definitions of
the volumetric mass transfer coefficient one can obtain the following concentration depending on
time, i.e. streamwise coordinate:
\beqal
&\vol t \frac{\ububble}{\ugas+\uliq}=\ln\frac{C^*}{C^*-C(t)}\\
&\vol \frac{\lunit}{\ugas+\uliq}=\frac{\lunit}{\ububble t} \ln \frac{C^*}{C^*-C(t)}\\
\feqal
Simulations for coefficient $\vol \frac{\ububble}{\ugas+\uliq}$ are shown in Fig.
\ref{fig:aver:conc:different:capillaries}.
\begin{figure}
\includegraphics[width=0.5\textwidth]{Figures/aver_conc_scale_ca026.eps}
\includegraphics[width=0.5\textwidth]{Figures/aver_conc_scale_ca054.eps}\\
\includegraphics[width=0.5\textwidth]{Figures/aver_conc_scale_ca05.eps}
\includegraphics[width=0.5\textwidth]{Figures/aver_conc_scale_ca097.eps}\\
\includegraphics[width=0.5\textwidth]{Figures/aver_conc_scale_ca14.eps}\\
\caption{Volumetric mass transfer coefficient for different capillaries and scales. One can see the
excellent agreement. \label{fig:aver:conc:different:capillaries}}
\end{figure}
One of interesting correlations for the mass transfer volumetric coefficient was presented by
\citet{yue-mass}:
\beqal
&\vol=\frac{2}{d_h}\Bigl(\frac{D \ububble}{\lbubble+\lslug}\Bigr)^{0.5}
\Bigl(\frac{\lbubble}{\lbubble+\lslug}\Bigr)^0.3\\
&\vol \frac{\lunit}{\ugas+\uliq}=2\frac{\lunit}{d_h} \Bigl(\frac{D \varepsilon
}{(\ububble+\ugas)\lbubble+\lslug}\Bigr)^{0.5}
\Bigl(\frac{\lbubble}{\lbubble+\lslug}\Bigr)^0.3 \propto Pe^{-\frac{1}{2}}.
\feqal
\begin{figure}
\includegraphics[width=\textwidth]{Figures/volumetric_mass_peclet.eps}
\end{figure}
